---
title: "Lab02: Causality Review"
date: "September 5, 2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Today's Goals

1. Garner more intuition

    - We want *causal effect*
    - Can't observe it for any given unit
    - *Randomization* helps us get at effect *on average*


## Today's Goals

2. Comfortably "speak the language" of randomized experiments

    - unit
    - treatment group/condition
    - control group/condition
    - outcome
    - potential outcomes
    - counterfactuals
    - causal effect
    - fundamental problem of causal inference
    - randomization
    - SATE
    - difference in means

## Today's Goals

3. A little R
 
 

## Running Example

- August 2006 Primary Statewide Election in Michigan
- Send postcards with randomly assigned message
    1. no message
    2. civic duty message
    3. “you are being studied” message 
    4. **neighborhood social pressure message**


- \footnotesize Source: Gerber, Alan S., Donald P. Green, and Christopher W. Larimer. "Social pressure and voter turnout: Evidence from a large-scale field experiment." American political Science review 102.1 (2008): 33-48.






## Discuss

- What is a **unit** in this study?
- What is the **outcome**?
- What is the **control group**?
- What are the **treatment groups**?

        
        
## Review

- What is a **unit** in this study?
    - a person
- What is the **outcome**?
    - whether or not the person voted, $Y$
- What is the **control group**?
    - subset of people who recieved no message
- What are the **treatment groups**?
    - subset of people who recieved
        - $T_1$: civic duty message
        - $T_2$: “you are being studied” message 
        - $T_3$: social pressure message

$\rightarrow$ Let's focus on social pressure message as $T$ of interest

## Discuss


- What are **potential outcomes**?
- What are the potential outcomes as applied to this study?
- How do you find the true **casual effect**?  Is this possible?
- What is the **fundamental problem of causal inference**?
- Explain the fundamental problem of causal inference as applied to this study.


## Review


- What are **potential outcomes**?
    - $Y(1)$ outcome if you recieved treatment
    - $Y(0)$ outcome if you recieved control
- What are the potential outcomes as applied to this study?
- How do you find the true **casual effect**?  Is this possible?
    - $Y(1) - Y(0)$
- What is the **fundamental problem of causal inference**?
    - can only observe one potential outcome
    - counterfactuals are not observed!
- Explain the fundamental problem of causal inference as applied to this study.


## Discuss

- Explain this equation in words: $SATE = \frac{1}{n} \sum_{i=1}^n \{Y_i(1) - Y_i(0)\}$ where $i$ indexes each unit
- Can we get an answer to this equation?  Why or why not?




## Review


- Explain this equation in words, where $i$ indexes each unit: $SATE = \frac{1}{n} \sum_{i=1}^n \{Y_i(1) - Y_i(0)\}$

- Can we get an answer to this equation? Why or why not?
    - no because counterfactuals aren't observed
    - i.e., the fundamental problem of causal inference!!
    - SATE is a theoretical concept so we must *estimate* it



## Discuss

- How do we do research given the fundamental problem of causal inference?
- What did this study do?
- How do we *estimate* SATE?



## Review

- How do we do research given the fundamental problem of causal inference?
    - Randomization
    - On average, treatment and control groups identical *but for which group they're assigned*
    - Compare *observed* outcomes across groups
- What did this study do?
    - Randomized, so groups were identical but for the message
    - Compare whether or not people voted across message groups
    - On average, was there a difference?
- How do we *estimate* SATE?
    - difference in means across groups


## Discuss

- Remember we must *estimate* SATE.
- We do it using the difference in means.
- Explain how this equation works:
    - diff in means = $\frac{1}{n_1} \sum_{i=1}^n T_i Y_i - \frac{1}{n-n_1} \sum_{i=1}^n (1-T_i)Y_i$
        - $n_1$ is the number of people in the treatment group
        - $T_i$ is an indicator s.t. $T_i = 0$ if unit recieved control and $T_i = 1$ if treatment


## Are you confortable "speaking the language" now?

- Hopefully this is getting repetative.
- If any questions remain...ask!


## A little R

- What is each row?
- Which variable indicates treament? 
- Which indicates outcome?
- What are the other variables for?

\footnotesize
```{r, echo = TRUE}
social <- read.csv("social.csv")
head(social)
```


## Estimate turnout rate (outcome) for each group

\footnotesize
```{r, echo = TRUE}
tapply(X = social$primary2006, INDEX = social$messages, FUN = mean)
```

```{r, echo = TRUE, eval = TRUE}
t1 <- mean(social$primary2006[social$messages == "Civic Duty"])
control <- mean(social$primary2006[social$messages == "Control"])
t2 <- mean(social$primary2006[social$messages == "Hawthorne"])
t3 <- mean(social$primary2006[social$messages == "Neighbors"])

c(t1, control, t2, t3)
```



## Estimate the SATE for each group

What do these results tell us?

\footnotesize
```{r, echo = TRUE, eval = TRUE}
t1 - control
t2 - control
t3 - control
```

## Randomization makes groups "identical" but for treatment assignment

\footnotesize
```{r, echo = TRUE, eval = TRUE}
tapply(social$primary2004, social$messages, mean)
tapply(social$hhsize, social$messages, mean)
social$sex <- ifelse(social$sex == "female", 1, 0)
tapply(social$sex, social$messages, mean)
```




