---
title: "Measurement and Plotting in R"
author: " "
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, root.dir = "~/Dropbox/Fall2018/TA/lab04/")
```


## Today's Goals

Review:

- interpretation/usefulness of different plots

- plotting syntax in R

- including plots in homework


## Reseach Question

Puzzle

- Does sharing a common gender unite or divide opposing partisans who identify as women?


Hypotheses:

1. **Opposing partisans who identify as women conceive of their gender identity differently.**

2. Mistrust increases among partisans who identify as women when the salience of gender increases.


Source

- Klar, Samara. 2018. “When Common Identities Decrease Trust: An Experimental Study of Partisan Women.” American Journal of Political Science. 62(3): 610-622.

## The Data

\footnotesize
```{r, echo = TRUE}
load("commonidentity.Rdata")
colnames(df)
```

  - `feminist` variables on 5 point scale: extremely well (1), very well (2), somewhat well (3), not very well (4), not at all (5)

  - To test hypothesis \#1, we'll focus on `feminist_describe_you` and `party_id` 



## Hypothesis #1

Survey to online panel of 3,607 women aged 18+ April 2017

What does this plot show?

```{r}
## you'll want to include code in your homework
tab_all <- table(df$feminist_describe_you)
barplot(tab_all,
        main = "Barplot of Identifying with Feminist Label",
        xlab = "Response",
        ylab = "Num. of Respondents",
        names.arg = c("extremely well", "very well", "somewhat well",
                      "not very well", "not at all", "NA"))
```


## Hypothesis #1

Does visual evidence support or refute hypothesis 1?


```{r}
tab_r <- table(df$feminist_describe_you[df$party_id == "Republican"])
tab_d <- table(df$feminist_describe_you[df$party_id == "Democrat"])
tab_both <- rbind(tab_r, tab_d)
barplot(tab_both,
        main = "Identifying with Feminist Label for Democrats (Blue) and Republicans (Red)",
        col = c("red", "blue"),
        xlab = "Response",
        ylab = "Num. of Respondents",
        names.arg = c("extremely well", "very well", "somewhat well",
                      "not very well", "not at all", "NA"))
```


## Alternative Hypothesis

The confounder of *age* explains both party identification and identifying with the "feminist" label.  Perhaps:

   - younger people more likely to be democrat

   - younger people more likely to identify as feminists
    
Let's examine with visuals.
    

## Visualizing Data: Histogram

- Histogram shows *density*
- What is density?
- Why does it sum to 100%?


## Visualizing Data: Histogram

What does this plot show?  Interpret first "box."

```{r}
h <- hist(x = df$age,
         freq = FALSE,
         breaks = 10,
         main = "Histogram of Age of Respondents",
         xlab = "Age")
h$density
areas <- 10 * h$density
sum(areas)
```



## Visualizing Data: Boxplots

What are the components of a boxplot?

What can we now say about our confounder suspicions?

```{r}
boxplot(df$age[df$party_id == "Republican"], df$age[df$party_id == "Democrat"],
        main = "Boxplots of Age by Party ID",
        names = c("Republican", "Democrat"),
        xlab = "Party ID",
        ylab = "Age")
```



## Visulaizing Data: Correlation

What is correlation?

  - Positive: when `x` increases, `y` increases
  - Negative: when `x` increases, `y` decreases
  - Strong vs. Weak
  - Between -1 and 1

Using different data so we have continuous variables.

  - Ask, does state spending on education correlate with student success?


## Visulaizing Data: Correlation

```{r, warning=FALSE, echo = TRUE, message=FALSE}
#install.packages("car")
library("car")
data("States")
?States ## look at variable descriptions
cor(x = States$dollars, y = States$SATV)
cor(x = States$dollars, y = States$SATM)
```
  
  
  
## Visulaizing Data: Scatterplot


```{r}
plot(x = States$dollars, y = States$SATV,
     xlab = "Spending on education in $1000s/student",
     ylab = "Average Verbal SAT Score",
     main = "Spending on Education and Student Success across States",
     cex = 1,
     col = "purple")
```



## Logarithmic Function

Useful to "smooth out" skewed variable:

- keep in mind for income, population, etc.

- functions in R for this

```{r, echo = TRUE}
log(3)
log(3, base = exp(1)) ## default base
log(3, base = 10)
```

## Logged Spending


```{r}
plot(x = log(States$dollars), y = States$SATV,
     xlab = "Log of spending on education in $1000s/student",
     ylab = "Average Verbal SAT Score",
     main = "Spending on Education and Student Success across States",
     cex = 1,
     col = "purple")
```

## Plotting in R markdown

To make plots look good in homwork pdf use r chunk options (see .Rmd).

```{r, fig.align = "center", fig.width = 3, fig.height = 3.5}
plot(x = 1, y = 1, xlab = "x", ylab = "y")
```



## Msc: Tips for Homework Success

- Lots and lots of plotting options... google questions! 
- Knit as you go.
- This homework is longer that's why you have 2 weeks...
- Remember, homework 3 we will start to deduct points for formatting if it impacts our ability to grade.  Questions?

    1. Write written answers to problems in the text space, not in comments in the R chunks.

    2. Relatedly, make sure lines of code and comments in R chunks do not run off the pdf page.

    3. Make sure that answers from code are displayed in the pdf.  If not working, use echo=T and eval=T in the r chunk options, explicitly.
