---
title: "Uncertainty and Hypothesis Tests"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Review

Why do we care about *uncertainty*?


## Review

Why do we care about *uncertainty*?

- to quantify how certain we can be of our empirical findings
- in other words, to quantify how likely is it that our result was just by chance

This is why we need probability theory.

## Review

In your own words, what's the main insight from each of these probability concepts?  Why are they important to political science research?


- LLN
- CLT


## Review

In your own words, what's the main insight from each of these probability concepts? Why are they important to political science research?

- LLN
    - as sample size increases, sample mean converges to population mean
- CLT
    - as sample size increases, sampling distribution of sample means is approximately normal 
    
    
    

    
## Today

- Testing hypotheses three ways:
    1. one sample test for mean
    2. two sample test for difference in means
    3. linear regression coefficient test

- Note the logic and framework is the same as we go through these tests








## Set up

Our research question: do female politicians promote different policies than men?

- Why might observation data not allow us to estimate a causal effect?


## Set up 


- Luckily for us, there was a randomized policy experiment in India since the mid 1990's in which 1/3 of village council heads have been *randomly* reserved for women.

- Why is this important? How does it allow us to estimate a causal effect of women as leaders on policies?

\vspace{2em}
\scriptsize
Source: Raghabendra Chattopadhyay and Esther Duflo. (2004). “Women as Policy Makers: Evidence from a Randomized Policy Experiment in India. *Econometrica*, Vol. 72, No. 5, pp. 1409-1443.


## Set up


West Bengal data:

- `water` measures number of new or repaired drinking water facilities in village since the reserve policy started
- `irrigation` measures the number of new or repaired irrigation facilities in the village since the reserve policy started
- `reserved` indicator for whether the Gram Panchayat (level of government of interest) was reserved for women leaders or not

\footnotesize
```{r}
link <- "https://raw.githubusercontent.com/kosukeimai/qss/master/PREDICTION/women.csv"
df <- read.csv(url(link), header = TRUE)
df[1:3, c("water", "irrigation", "reserved")]
```


## Hypothesis testing


Recall the steps for any hypothesis test (from the book, analagous to lecture):

1. State the null and alternative hypothesis.
2. Choose the appropriate test statistic and significance level $\alpha$.
3. Determine the sampling distribution of the test statistic *given the null is true*.
4. Compute the p-value.
3. Reject the null if p-value $\leq \alpha$, otherwise "retain" (fail to reject) the null.



## One sample test: Your task

An expert in this field thinks the average number of irrigation facilities across all the villages is 2.  Test the hypothesis the average is greater than 2. 

- Relevant info:

\footnotesize
```{r}
samp <- df$irrigation
mean(samp)
sd(samp)/sqrt(length(samp))
```




## One sample test: Answer

Step 1:

$$H_0: \mu = 2$$
$$H_A: \mu > 2$$


## One sample test: Answer

Step 2:

- We will use conventional level of significance of $\alpha = .05$

- We will use a Z-test statistic of the form:

$$Z = \frac{\bar{X} - \mu_0}{std. err. of \bar{X}}$$

- Remember standard error of mean is estimated by $\hat{sd(X)}/\sqrt{n}$

\footnotesize
```{r}
n <- length(df$irrigation)
std_err <- sd(df$irrigation)/sqrt(n)
z <- (mean(df$irrigation) - 2)/std_err
z #tip: z test-stat is unlikely to be bigger or smaller than, say, 4!
```



## One sample test: Answer


Step 3:

- By CLT we know $\bar{X} \sim N(\mu, \sigma/\sqrt{n})$
- and via "standardizing both sides"...$Z \sim N(0,1)$
- So given the null is true (given the mean number of irrigation facilities is no different than 4) Z is distributed $N(0,1)$

```{r, echo=FALSE, fig.asp=.5}
x_axis <- seq(-4,4,.01)
plot(x_axis, dnorm(x_axis), type = "l", main = "Sampling Distribution of Test Statistic",
     xlab = "Statistic Value", ylab = "Probability Given Null is True")
abline(v = z, col = "red")
text(x = z+.5, y = .05, labels = "z = 2.38")
```



## One sample test: Answer

Step 4:

- p-value is probability we'd observe this test statistic *or more extreme*, given the null is true

- Visually, does it seem unlikely we'd observe this result if the null were true?  Maybe... but probably not.  Let's calculate p-value.

```{r, echo=TRUE, fig.asp=.5}
pnorm(q = z, lower.tail = FALSE)
```






## One sample test: Answer

Step 5:

- $0.008 < .05$ so we reject the null (in favor of the alternative)


Note: important to know the steps, but R will do it for you
\footnotesize
```{r}
t.test(df$irrigation, mu = 2, alternative = "greater")
```





## Two sample test: Your task

We hypothesize that female politicians support different policies than men.  Test the hypothesis that reserving spots for female leaders influences the number of new/repaired drinking water facilities.

- Relevant info:

\footnotesize
```{r}
x_t <- df$water[df$reserved == 1] # outcome for treated units
x_c <- df$water[df$reserved == 0] # outcome for control units
mean(x_t)
mean(x_c)
std_error <- sqrt(var(x_t)/length(x_t) + var(x_c)/length(x_c))
std_error
```
























## Two Sample test: Answer

Step 1: 

- $H_0: \mu_T - \mu_C = 0$
- $H_A:  \mu_T - \mu_C \ne 0$


In words:

- Null--having reserved seats for female politicians *does not* influence the number of drinking water facilities in the villages.
- Alternative--having reserved seats for female politicians *does* influence the number of drinking water facilities in the villages.


## Two Sample test: Answer


Step 2:

- We will use conventional level of significance of $\alpha = .05$

- We will use a Z-test statistic of the form:

$$Z = \frac{(\bar{X_T } - \bar{X_C}) - 0}{\sqrt{\frac{1}{n_T} \hat{\sigma_T^2} + \frac{1}{n_C}\hat{\sigma_C^2}}}$$

- Remember standard error is different for different test stats

\footnotesize
```{r}
z <- (mean(x_t) - mean(x_c))/std_error
z
```


## Two sample test: Answer


Step 3:

- By CLT we know $Z \sim N(0,1)$
- So given the null is true (given the difference in means is 0) Z is distributed $N(0,1)$

```{r, echo=FALSE, fig.asp=.5}
x_axis <- seq(-4,4,.01)
plot(x_axis, dnorm(x_axis), type = "l", main = "Sampling Distribution of Test Statistic",
     xlab = "Statistic Value", ylab = "Probability Given Null is True")
abline(v = z, col = "red")
text(x = z+.5, y = .05, labels = "z = 1.81")
```




## One sample test: Answer

Step 4:

- Recall, p-value is probability we'd observe this test statistic *or more extreme*, given the null is true

- Visually, does it seem unlikely we'd observe this result if the null were true?  Maybe... maybe not?  Let's calculate p-value.

```{r, echo=TRUE, fig.asp=.5}
pnorm(q = -abs(z), lower.tail = TRUE)*2
```






## One sample test: Answer

Step 5:

- $0.0696 > .05$ so "retain" or fail to reject the null





## Regression

Forshadowing next week

- the steps for hypothesis testing for regression coefficients is no different
- by CLT,

$$ z-\text{ score of }\beta = \frac{\hat{\beta} - \beta}{\text{std. err. of }\hat{\beta}} \sim N(0,1)$$


Let's conduct the same hypothesis test using a regression framework.


## Regression: Answers

Step 1:

- $H_0: \beta = 0$
- $H_A: \beta \ne 0$



## Regression: Answer


Step 2: use conventional level of significance of $\alpha = .05$ and z-test statistic stated above


\footnotesize
```{r}
mod <- lm(water ~ reserved, data = df)
summary(mod)
```



## Regression: Answer


Step 3:

- By CLT we know $Z \sim N(0,1)$
- So given the null is true (given the coefficient is 0) Z is distributed $N(0,1)$

```{r, echo=FALSE, fig.asp=.5}
x_axis <- seq(-4,4,.01)
plot(x_axis, dnorm(x_axis), type = "l", main = "Sampling Distribution of Test Statistic",
     xlab = "Statistic Value", ylab = "Probability Given Null is True")
abline(v = 2.344, col = "red")
text(x = 2.344+.5, y = .05, labels = "z = 2.344")
```




## Regression: Answers


Step 4:

- Recall, p-value is probability we'd observe this test statistic *or more extreme*, given the null is true

- Visually, does it seem unlikely we'd observe this result if the null were true?  Probably! Let's calculate p-value.

```{r, echo=TRUE, fig.asp=.5}
pnorm(q = -abs(2.344), lower.tail = TRUE)*2
```











## Regression: Answer

Step 5:


- $0.0197 < \alpha$ at .05 level, so we reject the null

- So our finding that $\beta = 9.252$ was unlikely to be by chance.  The number of drinking facilities increased by about 9 due to the reservation of seats for women.






